Large and Infinitary Quotient Inductive-Inductive Types

### 2020-03-18 draft/paper.tex
cleanup references

\input{abbrevs.tex}
\newcommand{\Lift}{\Uparrow}
\newcommand{\ToS}{\mathsf{ToS}}
\newcommand{\ext}{\triangleright}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\Pii}{\Pi}
\newcommand{\appi}{\mathsf{app}}
\newcommand{\lami}{\mathsf{lam}}
\newcommand{\Pie}{\Pi^{\mathsf{ext}}}
\newcommand{\appe}{\mathsf{app^{ext}}}
\newcommand{\lame}{\mathsf{lam^{ext}}}
\newcommand{\Piinf}{\Pi^{\mathsf{inf}}}
\newcommand{\appinf}{\mathsf{app^{inf}}}
\newcommand{\laminf}{\mathsf{lam^{inf}}}
\newcommand{\appitt}{\mathop{{\scriptstyle @}}}
\newcommand{\Refl}{\mathsf{Refl}}
\newcommand{\IdU}{\mathsf{IdU}}
\newcommand{\ReflU}{\mathsf{ReflU}}
\newcommand{\Sig}{\mathsf{Sig}}
\newcommand{\ToSSig}{\mathsf{ToSSig}}
\newcommand{\Inductive}{\mathsf{Inductive}}
\newcommand{\Subtype}{\mathsf{Subtype}}
\newcommand{\subtype}{\mathsf{subtype}}
\newcommand{\NatSig}{\mathsf{NatSig}}

@@

Categories are a prime example for a structure which is described by a
quotient inductive-inductive signature. Signatures for QIITs allow
having multiple sorts, with later ones indexed over previous ones, and
equations as well. We need both features in order to write down the
following signature of categories.
\begingroup
\allowdisplaybreaks
\begin{alignat*}{3}
  & Ob  && : \U \\
  & Mor && : Ob\ra Ob\ra\U \\
  & \blank\circ\blank && : Mor\,J\,K\ra Mor\,I\,J\ra Mor\,I\,K \\
  & id && : Mor\,I\,I \\
  & ass && : (f\circ g)\circ h = f\circ(g\circ h) \\
  & idl && : id\circ f = f \\
  & idr && : f\circ id = f
\end{alignat*}
\endgroup
The benefit of having a QII signature is getting a model theory ``for free'',
from the metatheory of QIITs. This model theory includes a category of algebras
which has an initial object and also some additional structure. For the
signature of categories, we get the empty category as the initial object, but it
is common to consider categories with more structure, which have more
interesting initial models.

Algebraic notions of models of type theories are examples for this. Here,
initial models represent syntax, and initiality corresponds to induction on
syntax. Several variants have been used, from contextual categories \cite{gat}
and comprehension categories \cite{jacobs1993comprehension} to categories with
families \cite{Dybjer96internaltype} (CwF).

Whereas the initial category can be given by first defining the type of objects
(the empty type) and then defining morphisms separately, for an initial model of
a type theory given as a CwF one has to define $Ob$, $Mor$ and two extra sorts
$Ty$ and $Tm$ \emph{inductive-inductively}, at the same time. This is because
there is a constructor of $Ob$ which refers to $Ty$, constructors of $Ty$ refer
to $Ob$ and $Sub$, and so on. This is illustrated by the following excerpt of
the signature for CwFs extending the previous signature for categories.
\begin{alignat*}{3}
  & Ty && : Ob\ra \U \\
  & empty  && : Ob \\
  & ext && : (I:Ob)\ra Ty\,I\ra Ob \\
  & \blank[\blank] && : Ty\,J\ra Mor\,I\,J\ra Ty\,I
\end{alignat*}
While signatures for type theories usually only contain finitary constructors,
some QIITs have infinitary constructors, such as the following signature for
countably branching trees where the order of subtrees does not matter,
i.e.\ trees with permuted subtrees in a branch are equal.
\begin{alignat*}{3}
  & Tree && : \U \\
  & leaf  && : Tree \\
  & node && : (\N\ra Tree)\ra Tree \\
  & perm && : (g :\N\ra Tree)(f : \N\ra\N)\ra \mathsf{isIso}\,f\ra \\
  & && \hspace{0.8em} node\,g = node\,(g\circ f)
\end{alignat*}
Additional examples for infinitary QIITs are real, surreal numbers
\cite{HoTTbook}, ordinal numbers \cite{lumsdaineShulman} and the partiality
monad \cite{partiality}.

\subsection{Contributions}
The current paper extends the syntax and semantics of QIITs as previously described
in the literature \cite{kaposi2019constructing, gabe, dijkstra2017quotient}. We add the following features:
\begin{enumerate}
  \item
  \textbf{Large constructors, large elimination} and algebras at different
  universe levels. This fills in an important formal gap; large models are
  routinely used in the metatheory of type theories, but they have not been
  presented explicitly in previous QIIT literature.
  \item
  \textbf{Infinitary constructors}.
  Of special note here is that the theory of QIIT signatures is itself large and
  infinitary, thus it can ``eat itself'', i.e.\ include its own signature and
  provide its own metatheory. This was not possible previously in
  \cite{kaposi2019constructing}, where only finitary QIITs were described. In
  this paper we use self-representation to bootstrap the model theory of
  signatures, without having to assume any pre-existing internal syntax.
  \item
  \textbf{Recursive equations}, i.e. equations appearing as assumptions
  of constructors. These have occurred previously in syntaxes of cubical type
  theories, as boundary conditions \cite{cohen2016cubical, angiuli2016computational, angiuli2018cartesian}.
\end{enumerate}

@@

We have Russell-style predicative universes $\Set_i$ indexed by natural
numbers, dependent functions as $(x : A)\ra B$, and dependent pairs as $(x :
A)\times B$ with projections $\proj1$ and $\proj2$. We sometimes leave
parameters implicit in dependent function types, e.g.\ write $\id : A \ra A$
instead of $\id : (A : \Set_i)\ra A \ra A$. We also use subscripts as a field
projection notation for iterated pairs. For example, for $t : (A : \Set_i)
\times (B : \Set_i) \times (f : A \ra B)$, we use $B_t$ to denote the projection
of the second component. Sometimes we omit the subscript if it is clear from
context. When we write ``exists'' in this paper, we always mean chosen structure
given by a $\Sigma$-type.

@@

We use cumulativity to reduce bureaucratic overhead when dealing with
constructions at different universe levels. The internal $\Subtype$ is used in
Section \ref{sec:termmodels} to prove cumulativity for general QIIT
algebras. For example, consider natural number algebras at level $i$, given as
the $\Sigma$-type $NatAlg_i :=(Nat : \Set_i)\times Nat\times(Nat \ra
Nat)$. It follows from the subtyping rules that $i \leq j$ implies $NatAlg_{i}
\leq NatAlg_{j}$. However, cumulativity for arbitrary QIIT algebras does not
follow judgmentally; it can only be proven by induction on signatures, hence
the need for $\Subtype$.

@@

\textbf{Infinitary function space} $\Piinf : (A : \Set_j)\ra(A \ra
\Tm\,\Gamma\,\U)\ra \Tm\,\Gamma\,\U$, with $\appinf :
\Tm\,\Gamma\,(\El\,(\Pie\,A\,b))\ra((x : A)\ra \Tm\,\Gamma\,(\El\,(b\,x)))$ and
its inverse $\laminf$.

@@

\begin{example}{
    Signature for natural numbers. Here, no external types appear, so the level
    can be chosen as $0$.}
\begin{alignat*}{3}
  & \NatSig : \Sig_0 \\
  & \NatSig := \lambda (i :\mathbb{N})(M : \ToS_{i,0}).\\
  & \hspace{1em}(\emptycon_M\,\ext_M\, (N : \U_M) \,\ext_M\,(zero : \El_M\,N)\\
  & \hspace{2.67em}\ext_M (suc : N\arri_M\El_M\,N))
\end{alignat*}
\end{example}

@@

\begin{example}{Infinitary constructors}. The universe $\U$ is closed under
the $\Piinf$ function type, which allows such functions to appear in the domains
of $\Pii$ types. This allows, for example, a signature for trees branching with
arbitrary small types. This is a signature at level 1, since we have $\Set_0$ as
a $\Pie$ domain type.
\begin{alignat*}{3}
& \mathsf{TreeSig} := \\
& \hspace{1em}\emptycon \ext (Tree : \U)\\
& \hspace{1.35em}\ext (node : \Pie\,\Set_0\,(\lambda A.\,(A\,\arrinf\,Tree)\arri\El\,Tree))
\end{alignat*}
\end{example}

@@

\begin{example}{Recursive equations}. Again, the universe is closed under $\Id$, which allows us to write equations
in $\Pii$ domains. A minimal (and trivial) example:
\begin{alignat*}{3}
  & \mathsf{RecEqSig} :=\\
  & \hspace{0.5em}\emptycon\,\ext (A : \U) \ext (a : \El\,A)\ext (f : \Pii\,(x : A)\,(\Id\,A\,x\,a\arri\El\,A))
\end{alignat*}
More interesting (and complicated) examples for recursive equations are boundary
conditions in various cubical type theories \cite{cohen2016cubical, angiuli2016computational, angiuli2018cartesian}. Note that our $\Id$
allows iterated equations as well, but these are all trivial in the semantics,
where we assume UIP.
\end{example}

@@

Our semantics has a ``type-theoretic'' flavor, which is inspired by the cubical
set model of Martin-Löf type theory by Bezem et al. \cite{cubical}. The core idea
is to avoid strictness issues by starting from basic ingredients which are already
strict enough. Hence, instead of modeling types as certain slices and
substitution by pullback, we model types as displayed categories with extra
structure, which naturally support strict reindexing.

@@

In the following, for given $i$ and $j$ levels, we define a model $\bM_{i,j} :
\ToS_{\max(i+1,j)+1, j}$ such that $\Con_{\bM_{i,j}}$ is a type of structured
categories (of algebras). The level $i$ marks the level of all internal sorts in
an algebra, and the level $j$ marks the level of all external sets in function
domains. Hence, every algebra has level $\max(i+1,j)$. The bump is only needed
for $i$, since algebras merely contain elements of $A : \Set_j$ types, while
inductive sets are themselves elements of $\Set_i$. For example, $NatAlg_i :
\Set_{\max(i+1,0)} : \Set_{\max(i+1, 0)+1}$.

@@

A \emph{weak flCwF morphism} $\boldsymbol{\sigma : \Sub\,\Gamma\,\Delta}$ is a functor between underlying categories, which additionally maps types to types and terms to terms, and satisfies the following mere property:
  \begin{enumerate}
    \item $\bsigma\,(A[\sigma]) = (\bsigma\,A)\,[\bsigma\,\sigma]$
    \item $\bsigma\,(t[\sigma]) = (\bsigma\,t)\,[\bsigma\,\sigma]$
    \item The unique map $\epsilon : \Sub\,(\bsigma\,\emptycon)\,\emptycon$ has a retraction.
    \item Each $(\bsigma\,\p,\,\bsigma\,\q) : \Sub\,(\bsigma\,(\Gamma\,\ext\,A))\,(\bsigma\,\Gamma\,\ext\,\bsigma\,A)$ has an inverse.
  \end{enumerate}

In short, $\bsigma$ preserves substitution strictly and preserves empty context
and context extension up to isomorphism. We notate the evident isomorphisms as
$\bsigma_{\emptycon} : \bsigma\,\emptycon \simeq \emptycon$ and $\bsigma_{\ext}
: \bsigma\,(\Gamma\,\ext\,A)\,\simeq\,\bsigma\,\Gamma\,\ext\,\bsigma\,A$. Our
notion of weak morphism is the same as in \cite{birkedal2018modal}, when
restricted to CwFs.

@@

For $\boldsymbol{a : \Tm\,\Gamma\,\U}$ and $\boldsymbol{B :
  \Ty\,(\Gamma\ext\El\,a)}$, we aim to define $\boldsymbol{\Pi\,a\,B}
\boldsymbol{:} \bTy\,\bGamma$. We define this as a dependent product of
isofibrations, indexed by a discrete domain. The discreteness is essential: with
a general $\bA \boldsymbol{:} \bTy\,\bGamma$ domain, $\bPii$ would not be
definable because of variance issues. Indeed, the category of categories is not
locally cartesian closed and does not support a general $\Pi$ type
\cite{johnstone2002sketches}.

@@

We can convert a signature to a syntactic one by interpreting it in $\syn_j$,
and we can convert in the other direction by using ToS-induction to interpret a
$\Gamma : \Con_{\syn_j}$ in an arbitrary ToS model.

@@

In the following, We will need three model fragments of $\bM$, which can be used
to compute notions of algebras, displayed algebras and sections respectively for
each syntactic signature. This is a rephrasing of the $\blank^{\A}$,
$\blank^{\D}$ and $\blank^{\S}$ interpretations in
\cite{kaposi2019constructing}.

@@

\begin{definition}[Logical predicate model of $\ToS$ over the $\Set$ model]
For each $i$ and $j$ level we have $\bD$, which is a displayed $\ToS$ model over
$\bA$. This model, analogously to $\bA$, is given by restricting $\bM_{i,j}$ to
the $\Ty$ components everywhere, corresponding to types or actions on
types. Hence, we have:
\begin{alignat*}{3}
  & \Con_{\bD}\,\ulGamma && = \ulGamma \ra \Set_{\max(i+1, j)} \\
  & \Sub_{\bD}\,\Gamma\,\Delta\,\ulsigma && = (\ulgamma : \ulGamma) \ra \Gamma\,\gamma \ra \Delta\,(\ulsigma\,\ulgamma)\\
  & \Ty_{\bD}\,\Gamma\,\ulA && = (\ulgamma : \ulGamma) \ra \Gamma\,\ulgamma \ra \ulA\,\ulgamma \ra \Set_{\max(i+1, j)}\\
  & \Tm_{\bD}\,\Gamma\,A\,\ult && = (\ulgamma : \ulGamma)(\gamma : \Gamma\,\ulGamma)
    \ra A\,\ulgamma\,\gamma\,(\ult\,\ulgamma)
\end{alignat*}
\end{definition}

@@

The basic idea is that initial algebras can be built from the terms of $\syn_j$.
For example, consider the syntactic signature for natural numbers:
\[
\NatSig := \emptycon\ext(N : \U)\ext(zero : \El\,N)\ext (suc : N
\arri \El\,N)
\]
The type $\Tm_{\syn}\,\NatSig\,(\El_{\syn}\,N)$ is isomorphic to the usual type
of natural numbers, since, intuitively, such terms can only be built from
iterated usage of $zero$ and $suc$. We build a term algebra for each
signature in this manner.

\begin{definition}[Term algebra construction]
For each syntactic signature $\ulOmega : \SynSig_j$, we define a displayed $\ToS$
model over $\syn_j$, named $\bT_{\ulOmega}$. The underlying sets are as follows:
\begin{alignat*}{3}
  & \Con_{\bT_{\ulOmega}}\,\ulGamma &&:= \Sub\,\ulOmega\,\ulGamma \ra \llb\ulGamma\rrb_{\bA}\\
  & \Sub_{\bT_{\ulOmega}}\,\Gamma\,\Delta\,\ulsigma &&:= (\ulnu : \Sub\,\ulOmega\,\ulGamma)
  \ra \Delta\,(\ulsigma\circ\nu) \simeq \llb\ulsigma\rrb_{\bA}\,(\Gamma\,\ulnu)\\
  & \Ty_{\bT_{\ulOmega}}\,\Gamma\,\ulA &&:= (\ulnu : \Sub\,\ulOmega\,\ulGamma) \ra
    \Tm\,\ulOmega\,(\ulA[\ulnu]) \ra \llb\ulA\rrb_{\bA}\,(\Gamma\,\ulnu)\\
  & \Tm_{\bT_{\ulOmega}}\,\Gamma\,A\,\ult &&:= (\ulnu : \Sub\,\ulOmega\,\ulGamma)\ra
    \llb\ulA\rrb_{\bA}\,\ulnu\,(\ult[\ulnu]) \simeq_{\id} \llb\ult\rrb_{\bA} (\Gamma\,\ulnu)
\end{alignat*}

@@

Higher inductive types (HITs) are generalizations of QIITs in settings with
proof-relevant identity types. They were introduced before QIITs
\cite{HoTTbook}. \cite{niels} is a description of a subset of HITs. \cite{hiit}
describes a syntax for higher inductive-inductive types using a theory of
signatures similar to ours, however does not construct categories of algebras or
show initiality. Semantics for different subclasses of HITs are given by
\cite{lumsdaineShulman,10.1145/3209108.3209130,nielsmsc,sojakova,moeneclaey}. Cubical
type theories were shown to support HITs in a computational way
\cite{niels,hiit}.

@@

With regards to the reduction of QIITs to simple type formers, the
reduction of infinitary QIITs appears to be more challenging than the finitary
case. \cite[Section 9]{lumsdaineShulman} shows that infinitary QIITs are not
constructible in type theory without choice, so such QIITs are a
non-conservative extension of extensional type theory. In the finitary case, a
generalization of the approach in \cite{induction-is-enough} seems promising ;
this amounts to a Streicher-style initial algebra construction
\cite{streicher2012semantics} for the theory of finitary QIIT signatures. In
particular, Brunerie et al. \cite{brunerie} has formalized in Agda this
construction for a comparable type theory, using UIP, function extensionality,
propositional extensionality and simple quotient types.

>>>>

\input{abbrevs.tex}
\newcommand{\Lift}{\Uparrow}
\newcommand{\ToS}{\mathsf{ToS}}
\newcommand{\ext}{\triangleright}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\Pii}{\Pi}
\newcommand{\appi}{\mathsf{app}}
\newcommand{\lami}{\mathsf{lam}}
\newcommand{\Pie}{\Pi^{\mathsf{ext}}}
\newcommand{\appe}{\mathsf{app^{ext}}}
\newcommand{\lame}{\mathsf{lam^{ext}}}
\newcommand{\Piinf}{\Pi^{\mathsf{inf}}}
\newcommand{\appinf}{\mathsf{app^{inf}}}
\newcommand{\laminf}{\mathsf{lam^{inf}}}
\newcommand{\appitt}{\mathop{{\scriptstyle @}}}
\newcommand{\Refl}{\mathsf{Refl}}
\newcommand{\IdU}{\mathsf{IdU}}
\newcommand{\ReflU}{\mathsf{ReflU}}
\newcommand{\Sig}{\mathsf{Sig}}
\newcommand{\ToSSig}{\mathsf{ToSSig}}
\newcommand{\Inductive}{\mathsf{Inductive}}
\newcommand{\Subtype}{\mathsf{Subtype}}
\newcommand{\subtype}{\mathsf{subtype}}
\newcommand{\NatSig}{\mathsf{NatSig}}
\newcommand{\mi}[1]{\mathit{#1}}

@@

Categories are a prime example for a structure which is described by a
quotient inductive-inductive signature. Signatures for QIITs allow
having multiple sorts, with later ones indexed over previous ones, and
equations as well. We need both features in order to write down the
following signature of categories.
\begingroup
\allowdisplaybreaks
\begin{alignat*}{3}
  & \mi{Ob}  && : \U \\
  & \mi{Mor} && : \mi{Ob}\ra \mi{Ob}\ra\U \\
  & \blank\circ\blank && : \mi{Mor}\,J\,K\ra \mi{Mor}\,I\,J\ra \mi{Mor}\,I\,K \\
  & \mi{id} && : \mi{Mor}\,I\,I \\
  & \mi{ass} && : (f\circ g)\circ h = f\circ(g\circ h) \\
  & \mi{idl} && : \mi{id}\circ f = f \\
  & \mi{idr} && : f\circ \mi{id} = f
\end{alignat*}
\endgroup
The benefit of having a QII signature is getting a model theory ``for free'',
from the metatheory of QIITs. This model theory includes a category of algebras
which has an initial object and also some additional structure. For the
signature of categories, we get the empty category as the initial object, but it
is common to consider categories with more structure, which have more
interesting initial models.

Algebraic notions of models of type theories are examples for this. Here,
initial models represent syntax, and initiality corresponds to induction on
syntax. Several variants have been used, from contextual categories \cite{gat}
and comprehension categories \cite{jacobs1993comprehension} to categories with
families \cite{internaltt} (CwF).

Whereas the initial category can be given by first defining the type of objects
(the empty type) and then defining morphisms separately, for an initial model of
a type theory given as a CwF one has to define $\mi{Ob}$, $\mi{Mor}$ and two
extra sorts $\mi{Ty}$ and $\mi{Tm}$ \emph{inductive-inductively}, at the same
time. This is because there is a constructor of $\mi{Ob}$ which refers to
$\mi{Ty}$, constructors of $\mi{Ty}$ refer to $\mi{Ob}$ and $\mi{Sub}$, and so
on. This is illustrated by the following excerpt of the signature for CwFs
extending the previous signature for categories.
\begin{alignat*}{3}
  & \mi{Ty}        && : \mi{Ob}\ra \U \\
  & \mi{empty}     && : \mi{Ob} \\
  & \mi{ext}       && : (I:\mi{Ob})\ra \mi{Ty}\,I\ra \mi{Ob} \\
  & \blank[\blank] && : \mi{Ty}\,J\ra \mi{Mor}\,I\,J\ra \mi{Ty}\,I
\end{alignat*}
While signatures for type theories usually only contain finitary constructors,
some QIITs have infinitary constructors, such as the following signature for
countably branching trees where the order of subtrees does not matter,
i.e.\ trees with permuted subtrees in a branch are equal.
\begin{alignat*}{3}
  & \mi{Tree} && : \U \\
  & \mi{leaf}  && : \mi{Tree} \\
  & \mi{node} && : (\N\ra \mi{Tree})\ra \mi{Tree} \\
  & \mi{perm} && : (g :\N\ra \mi{Tree})(f : \N\ra\N)\ra \mathsf{isIso}\,f\ra \\
  & && \hspace{0.8em} \mi{node}\,g = \mi{node}\,(g\circ f)
\end{alignat*}
Additional examples for infinitary QIITs are real, surreal numbers
\cite{HoTTbook}, ordinal numbers \cite{lumsdaineShulman} and the partiality
monad \cite{partiality}.

\subsection{Contributions}
The current paper extends the syntax and semantics of QIITs as previously described
in the literature \cite{kaposi2019constructing, qiits, dijkstra2017quotient}. We add the following features:
\begin{enumerate}
  \item
  \textbf{Large constructors, large elimination} and algebras at different
  universe levels. This fills in an important formal gap; large models are
  routinely used in the metatheory of type theories, but they have not been
  presented explicitly in previous QIIT literature.
  \item
  \textbf{Infinitary constructors}.
  Of special note here is that the theory of QIIT signatures is itself large and
  infinitary, thus it can ``eat itself'', i.e.\ include its own signature and
  provide its own metatheory. This was not possible previously in
  \cite{kaposi2019constructing}, where only finitary QIITs were described. In
  this paper we use self-representation to bootstrap the model theory of
  signatures, without having to assume any pre-existing internal syntax.
  \item
  \textbf{Recursive equations}, i.e. equations appearing as assumptions
  of constructors. These have occurred previously in syntaxes of cubical type
  theories, as boundary conditions \cite{cchm, angiuli2016computational, angiuli2018cartesian}.
\end{enumerate}

@@

We have Russell-style predicative universes $\Set_i$ indexed by natural numbers,
dependent functions as $(x : A)\ra B$, and dependent pairs as $(x : A)\times
B$. We sometimes leave parameters implicit in dependent function types,
e.g.\ write $\id : A \ra A$ instead of $\id : (A : \Set_i)\ra A \ra A$. We also
use subscripts as a field projection notation for iterated pairs. For example,
for $t : (A : \Set_i) \times (B : \Set_i) \times (f : A \ra B)$, we use $B_t$ to
denote the projection of the second component. Sometimes we omit the subscript
if it is clear from context. When we write ``exists'' in this paper, we always
mean chosen structure given by a $\Sigma$-type.

@@

We use cumulativity to reduce bureaucratic overhead when dealing with
constructions at different universe levels. The internal $\Subtype$ is used in
Section \ref{sec:termmodels} to prove cumulativity for general QIIT
algebras. For example, consider natural number algebras at level $i$, given as
the $\Sigma$-type $\mi{NatAlg}_i :=(\mi{Nat} : \Set_i)\times
\mi{Nat}\times(\mi{Nat} \ra \mi{Nat})$. It follows from the subtyping rules that
$i \leq j$ implies $NatAlg_{i} \leq \mi{NatAlg}_{j}$. However, cumulativity for
arbitrary QIIT algebras does not follow judgmentally; it can only be proven by
induction on signatures, hence the need for $\Subtype$.

@@

\textbf{Infinitary function space} $\Piinf : (A : \Set_j)\ra(A \ra
\Tm\,\Gamma\,\U)\ra \Tm\,\Gamma\,\U$, with $\appinf :
\Tm\,\Gamma\,(\El\,(\Piinf\,A\,b))\ra((x : A)\ra \Tm\,\Gamma\,(\El\,(b\,x)))$ and
its inverse $\laminf$.

@@

\begin{example}{
    Signature for natural numbers. Here, no external types appear, so the level
    can be chosen as $0$.}
\begin{alignat*}{3}
  & \NatSig : \Sig_0 \\
  & \NatSig := \lambda (i :\mathbb{N})(M : \ToS_{i,0}).\\
  & \hspace{1em}(\emptycon_M\,\ext_M\, (N : \U_M) \,\ext_M\,(\mi{zero} : \El_M\,N)\\
  & \hspace{2.67em}\ext_M (\mi{suc} : N\arri_M\El_M\,N))
\end{alignat*}
\end{example}

@@

\begin{example}{Infinitary constructors}. The universe $\U$ is closed under
the $\Piinf$ function type, which allows such functions to appear in the domains
of $\Pii$ types. This allows, for example, a signature for trees branching with
arbitrary small types. This is a signature at level 1, since we have $\Set_0$ as
a $\Pie$ domain type.
\begin{alignat*}{3}
& \mathsf{TreeSig} := \\
& \hspace{1em}\emptycon \ext (\mi{Tree} : \U)\\
& \hspace{1.35em}\ext (\mi{node} : \Pie\,\Set_0\,(\lambda A.\,(A\,\arrinf\,\mi{Tree})\arri\El\,\mi{Tree}))
\end{alignat*}
\end{example}

@@

\begin{example}{Recursive equations}. Again, the universe is closed under $\Id$, which allows us to write equations
in $\Pii$ domains. A minimal (and trivial) example:
\begin{alignat*}{3}
  & \mathsf{RecEqSig} :=\\
  & \hspace{0.5em}\emptycon\,\ext (A : \U) \ext (a : \El\,A)\ext (f : \Pii\,(x : A)\,(\Id\,A\,x\,a\arri\El\,A))
\end{alignat*}
More interesting (and complicated) examples for recursive equations are boundary
conditions in various cubical type theories \cite{cchm, angiuli2016computational, angiuli2018cartesian}. Note that our $\Id$
allows iterated equations as well, but these are all trivial in the semantics,
where we assume UIP.
\end{example}

@@

Our semantics has a ``type-theoretic'' flavor, which is inspired by the cubical
set model of Martin-Löf type theory by Bezem et al. \cite{cubicalmodel}. The core idea
is to avoid strictness issues by starting from basic ingredients which are already
strict enough. Hence, instead of modeling types as certain slices and
substitution by pullback, we model types as displayed categories with extra
structure, which naturally support strict reindexing.

@@

In the following, for given $i$ and $j$ levels, we define a model $\bM_{i,j} :
\ToS_{\max(i+1,j)+1, j}$ such that $\Con_{\bM_{i,j}}$ is a type of structured
categories (of algebras). The level $i$ marks the level of all internal sorts in
an algebra, and the level $j$ marks the level of all external sets in function
domains. Hence, every algebra has level $\max(i+1,j)$. The bump is only needed
for $i$, since algebras merely contain elements of $A : \Set_j$ types, while
inductive sets are themselves elements of $\Set_i$. For example, $\mi{NatAlg}_i :
\Set_{\max(i+1,0)} : \Set_{\max(i+1, 0)+1}$.

@@

A \emph{weak flCwF morphism} $\boldsymbol{\sigma : \Sub\,\Gamma\,\Delta}$ is a
functor between underlying categories, which also maps types to types and terms
to terms, and satisfies the following mere properties:
  \begin{enumerate}
    \item $\bsigma\,(A[\sigma]) = (\bsigma\,A)\,[\bsigma\,\sigma]$
    \item $\bsigma\,(t[\sigma]) = (\bsigma\,t)\,[\bsigma\,\sigma]$
    \item The unique map $\epsilon : \Sub\,(\bsigma\,\emptycon)\,\emptycon$ has a retraction.
    \item Each $(\bsigma\,\p,\,\bsigma\,\q) : \Sub\,(\bsigma\,(\Gamma\,\ext\,A))\,(\bsigma\,\Gamma\,\ext\,\bsigma\,A)$ has an inverse.
  \end{enumerate}

In short, $\bsigma$ preserves substitution strictly and preserves empty context
and context extension up to isomorphism. We notate the evident isomorphisms as
$\bsigma_{\emptycon} : \bsigma\,\emptycon \simeq \emptycon$ and $\bsigma_{\ext}
: \bsigma\,(\Gamma\,\ext\,A)\,\simeq\,\bsigma\,\Gamma\,\ext\,\bsigma\,A$. Our
notion of weak morphism is the same as in \cite{birkedal2020modal}, when
restricted to CwFs.

@@

For $\boldsymbol{a : \Tm\,\Gamma\,\U}$ and $\boldsymbol{B :
  \Ty\,(\Gamma\ext\El\,a)}$, we aim to define $\boldsymbol{\Pi\,a\,B}
\boldsymbol{:} \bTy\,\bGamma$. We define this as a dependent product of
isofibrations, indexed by a discrete domain. The discreteness is essential: with
a general $\bA \boldsymbol{:} \bTy\,\bGamma$ domain, $\bPii$ would not be
definable because of variance issues. Indeed, the category of categories is not
locally cartesian closed and does not support a general $\Pi$ type
\cite[Section A1.5]{johnstone2002sketches}.

@@

We can convert a signature to a syntactic one by interpreting it in $\syn_j$,
and we can convert in the other direction by using ToS-induction to interpret a
$\Gamma : \Con_{\syn_j}$ in an arbitrary ToS model. This is merely a logical
equivalence, external to cETT (because of unvierse polymorphism), and not an
isomorphism.

@@

In the following, we will need three model fragments of $\bM$, which can be used
to compute notions of algebras, displayed algebras and sections respectively for
each syntactic signature. This is a rephrasing of the $\blank^{\A}$,
$\blank^{\D}$ and $\blank^{\S}$ interpretations in
\cite{kaposi2019constructing}.

@@

\begin{definition}[Logical predicate model of $\ToS$ over the $\Set$ model]
For each $i$ and $j$ level we have $\bD$, which is a displayed $\ToS$ model over
$\bA$. This model, analogously to $\bA$, is given by restricting $\bM_{i,j}$ to
the $\Ty$ components everywhere, corresponding to types or actions on
types. Hence, we have:
\begin{alignat*}{3}
  & \Con_{\bD}\,\ulGamma && = \ulGamma \ra \Set_{\max(i+1, j)} \\
  & \Sub_{\bD}\,\Gamma\,\Delta\,\ulsigma && = (\ulgamma : \ulGamma) \ra \Gamma\,\ulgamma \ra \Delta\,(\ulsigma\,\ulgamma)\\
  & \Ty_{\bD}\,\Gamma\,\ulA && = (\ulgamma : \ulGamma) \ra \Gamma\,\ulgamma \ra \ulA\,\ulgamma \ra \Set_{\max(i+1, j)}\\
  & \Tm_{\bD}\,\Gamma\,A\,\ult && = (\ulgamma : \ulGamma)(\gamma : \Gamma\,\ulgamma)
    \ra A\,\ulgamma\,\gamma\,(\ult\,\ulgamma)
\end{alignat*}
\end{definition}

@@

The basic idea is that initial algebras can be built from the terms of $\syn_j$.
For example, consider the syntactic signature for natural numbers:
\[
\NatSig := \emptycon\ext(N : \U)\ext(\mi{zero} : \El\,N)\ext (\mi{suc} : N
\arri \El\,N)
\]
The type $\Tm_{\syn}\,\NatSig\,(\El_{\syn}\,N)$ is isomorphic to the usual type
of natural numbers, since, intuitively, such terms can only be built from
iterated usage of $\mi{zero}$ and $\mi{suc}$. We build a term algebra for each
signature in this manner.

\begin{definition}[Term algebra construction]
For each syntactic signature $\ulOmega : \SynSig_j$, we define a displayed $\ToS$
model over $\syn_j$, named $\bT_{\ulOmega}$. The underlying sets are as follows:
\begin{alignat*}{3}
  & \Con_{\bT_{\ulOmega}}\,\ulGamma &&:= \Sub\,\ulOmega\,\ulGamma \ra \llb\ulGamma\rrb_{\bA}\\
  & \Sub_{\bT_{\ulOmega}}\,\Gamma\,\Delta\,\ulsigma &&:= (\ulnu : \Sub\,\ulOmega\,\ulGamma)
  \ra \Delta\,(\ulsigma\circ\ulnu) \simeq \llb\ulsigma\rrb_{\bA}\,(\Gamma\,\ulnu)\\
  & \Ty_{\bT_{\ulOmega}}\,\Gamma\,\ulA &&:= (\ulnu : \Sub\,\ulOmega\,\ulGamma) \ra
    \Tm\,\ulOmega\,(\ulA[\ulnu]) \ra \llb\ulA\rrb_{\bA}\,(\Gamma\,\ulnu)\\
  & \Tm_{\bT_{\ulOmega}}\,\Gamma\,A\,\ult &&:= (\ulnu : \Sub\,\ulOmega\,\ulGamma)\ra
    \llb\ulA\rrb_{\bA}\,\ulnu\,(\ult[\ulnu]) \simeq_{\id} \llb\ult\rrb_{\bA} (\Gamma\,\ulnu)
\end{alignat*}

@@

Higher inductive types (HITs) are generalizations of QIITs in settings with
proof-relevant identity types. They were introduced before QIITs
\cite{HoTTbook}. \cite{hitsinprogramming} is a description of a subset of HITs. \cite{hiit}
describes a syntax for higher inductive-inductive types using a theory of
signatures similar to ours, however does not construct categories of algebras or
show initiality. Semantics for different subclasses of HITs are given by
\cite{lumsdaineShulman,10.1145/3209108.3209130,nielsmsc,sojakova,moeneclaey}. Cubical
type theories were shown to support HITs in a computational way
\cite{niels,hiit}.

@@

With regards to the reduction of QIITs to simple type formers, the reduction of
infinitary QIITs appears to be more challenging than the finitary
case. \cite[Section 9]{lumsdaineShulman} shows that infinitary QIITs are not
constructible from inductive types and simple quotients with relations. In the
finitary case, a generalization of the approach in \cite{induction-is-enough}
seems promising ; this amounts to a Streicher-style initial algebra construction
\cite{streicher2012semantics} for the theory of finitary QIIT signatures. In
particular, Brunerie et al. \cite{brunerie} have formalized in Agda this
construction for a comparable type theory, using UIP, function extensionality,
propositional extensionality and simple quotient types.
